---
title: "Graphs R Us - Data Visualisation with R Project"
author: "Pawaneet Kaur"
output: rmarkdown::github_document

---


## setting up work directory: 

getwd ()Indicates the current working directory without no arguments by returning a NULL or character string. # This is important for debugging, larger programs to reiterate via the directory contents. 

On Windows the returned path use "/" between directory levels as a separator unless in a root directory (of a share or drive on Windows)

setwd() function changes and sets the existing working director navigated via getwd function.



## chilling_sensitivity.csv


```{r}
cs = read.csv("chilling_sensitivity.csv") # activating the chilling_sensitivity.csv dataframe; represented  as "cs"
```



## drought_elevation.csv


```{r}
elv = read.csv("drought_elevation.csv") # activating the drought_elevation.csv dataframe; displayed as "elv"

```



The "read.csv" reads a excel table format files to generate a data frame 




### Part 1 ###



##  Data Aggregation 


```{r}
library(dplyr)

```


```{r}
head(cs, n=3) # the x axis independent variable is ambient CO2 concentration (ppm) 
#the  while the y axis dependent variable is the  photosynthetic rate (µmol m-2 s -1)
```



# viewing original data via scatterplot

```{r}
plot( photo ~ conc, data = cs)

```


# Standard error function

```{r}
se <- function(x, na.rm = FALSE)# a.rm = FALSE removes blank cells 
{
  if(na.rm == TRUE)
  {
    sqrt(var(x, na.rm = T)/length(na.omit(x)))
  }
  else
  {
    sqrt(var(x)/length(x))
  }
}


pcs <- group_by(.data = cs, treat, spec,conc) %>% summarise(p_avv = mean(photo)
    , pos = mean(photo) + se(photo), neg = mean(photo) - se(photo))
head(pcs,n=3)

```


# without concentration 

```{r,warning=FALSE,message=FALSE,results='hide'}
pcsnc <- group_by(.data = cs, treat, spec) %>% summarise(p_avvnc = mean(photo)
, pos = mean(photo) + se(photo), neg = mean(photo) - se(photo))
```


## Wide formating our photosynthesis rate data

```{r,warning=FALSE,message=FALSE,results='hide'}
library(reshape2) # required to run acast function
wides= acast(data = pcsnc, formula = treat ~ spec, value.var = "p_avvnc")
wides
```



##  Multi-panel scatterplot in traditional graphics system 

I have not subsetted from the dplyr package. 


```{r}
# four panel Layout

par(mfrow=c(2, 2)) # 2 rows and 2 columns

# Dots/points in blue indicate nonchilled and red as chilled 

##### species #####

## Lolium perenne 

plot(cs$photo[cs$spec=="Lolium perenne"][cs$treat=="nonchilled"] ~ 
       cs$conc[cs$spec=="Lolium perenne"] [cs$treat=="nonchilled"]
     , col= "blue", pch= 17, ylab= expression(paste("Photosynthesis rate ( ", mu," mol m"^-2, " s"^-1, ")"))
     ,xlab='', main = substitute(paste(italic('Lolium perenne'))))
points(cs$photo[cs$spec=="Lolium perenne"][cs$treat=="chilled"] ~ 
         cs$conc[cs$spec=="Lolium perenne"][cs$treat=="chilled"]
       , col= "red",pch= 20)

# legend

l <- legend( "bottomright"
             , inset = c(0,0) 
             , cex = 1
             , bty = "n"
             , legend = c("Chilled", "Non Chilled")
             , text.col = c("red", "blue")
             , pt.bg = c("red","blue")
             , pch = c(20,17)
             , col = c("red","blue"))



## Cynosurus cristatus 

plot(cs$photo[cs$spec=="Cynosurus cristatus"] [cs$treat=="nonchilled"] ~ 
       cs$conc[cs$spec=="Cynosurus cristatus"] [cs$treat=="nonchilled"]
     , col= "blue", pch= 17, ylab='',xlab='', main = substitute(paste(italic('Cynosurus cristatus')))) 

points(cs$photo[cs$spec=="Cynosurus cristatus"][cs$treat=="chilled"] ~ 
         cs$conc[cs$spec=="Cynosurus cristatus"][cs$treat=="chilled"]
       , col= "red",pch= 20)


## Dactylis glomerata  

plot(cs$photo[cs$spec=="Dactylis glomerata"] [cs$treat=="nonchilled"] ~ 
       cs$conc[cs$spec=="Dactylis glomerata"] [cs$treat=="nonchilled"]
     , col= "blue", pch= 17, ylab= expression(paste("Photosynthesis rate ( ",  mu," mol m"^-2, " s"^-1, ")"))
     ,xlab= expression(paste("Ambient CO"[2], " concentration (ppm)")), main = substitute(paste(italic('Dactylis glomerata'))))
points(cs$photo[cs$spec=="Dactylis glomerata"] [cs$treat=="chilled"] ~ 
         cs$conc[cs$spec=="Dactylis glomerata"] [cs$treat=="chilled"]
       , col= "red",pch= 20)


## Holcus mollis 

plot(cs$photo[cs$spec=="Holcus mollis"] [cs$treat=="nonchilled"] ~ 
       cs$conc[cs$spec=="Holcus mollis"] [cs$treat=="nonchilled"]
     , col= "blue", pch= 17, ylab='',xlab= expression(paste("Ambient CO"[2], " concentration (ppm)")), main = substitute(paste(italic('Holcus mollis'))))
points(cs$photo[cs$spec=="Holcus mollis"] [cs$treat=="chilled"] ~ 
         cs$conc[cs$spec=="Holcus mollis"] [cs$treat=="chilled"]
       , col= "red",pch= 20)



# Note: the species, co2 concentation (xlab= ) and Photosynthesis (ylab= ) was named within plot () function.
 
```


#  Multi-panel scatterplot in ggplot2 

 Multi-panel scatterplot in ggplot2 
 
```{r}
library(ggplot2) 
library(dplyr)
```

Photosynthesis rate is higher on non chilled treatment

```{r}
qplot(conc,photo, data = cs, color= treat, facets=.~ spec) + labs(x = expression(paste("Ambient CO"[2], " concentration (ppm)"))
    , y= expression(paste("Photosynthesis rate ( ",  mu,"mol m"^-2, " s"^-1, ")"))) +
  theme_light() + 
  scale_colour_manual(labels= c("Chilled", "Non-chilled"), values= c("red","blue")) +
  labs(colour="Treatment") +
  theme(strip.text = element_text(face = "italic")) + 
  theme(legend.title =element_text(size=10)) + 
  theme(axis.title.x = element_text(size=10)) + 
  theme(axis.title.y = element_text(size=10))
```






### Part 2 ###


## Multi-panel barplot with insets 



```{r}
# Generating  aggregating and dataframe  data

de =read.csv("drought_elevation.csv")
head(de)
str(de)
```



```{r}
## Creating standard error (se) function

se <- function(x, na.rm = FALSE)
{
  if(na.rm == TRUE)
  {
    sqrt(var(x, na.rm = T)/length(na.omit(x)))
  }
  else
  {
    sqrt(var(x)/length(x))
  }
}
```



Generating the Multi-panel barplot

```{r,warning=FALSE,message=FALSE,results='hide'}

##  insets and multipanel layout 

layout(widths = c(0.9, 0.3),rbind(c(1, 2),
             c(1, 1),
             c(3, 4),
             c(3, 3),
             c(5, 6),
             c(5, 5)))
layout.show(n = 6)

oper <- par(oma = c(1,1,1,1),mar = c(1,1,1,1))

## top panel ##

# Aggregating high elevation 

de4 <- de[!de$elevation %in% c("mid","low"), ]
de4 <- droplevels(de4) 
levels(de4$elevation) # selecting high elevation
de5 <- group_by(.data = de4, spec, treat ) %>%
  summarise(vertical_mean = mean(height, na.rm = T),
            neg = mean(height, na.rm = T) - se(height, na.rm = T),
             pos= mean(height, na.rm = T) + se(height, na.rm = T))
de5 <- de5[order(de5$treat), ]

# Panel layout

par(mar = c(.24, 2.1, .95, .55))

# Wide format data transformation 

format_wide <- acast(data= de5, treat ~ spec, value.var = "vertical_mean")

# High elevation barplot

barplot= barplot(format_wide, col = c("white", "blue"), ylim = c(0, 60), axisnames = F,  axes = T, 
  beside = T) # barplot keeps the midpoints bars coordinates; creating texts/error bars 
##--------- ---------##------ ------##
## Extra 

# legend
legend(x = 0.8, y = 58, legend = c("Drought","Control"), bty = "n",
       pch = 22,
       pt.bg = c("blue", "white"), pt.cex = 2, cex = 1, y.intersp =
         1.1, xpd = NA)
##--------- ---------##------ ------##

# High elevation title

mtext("High", outer = F, line = -1.7,side = 3, cex = 0.8)

# Error bars

arrows(x0 = barplot,x1 = barplot, y1 = de5$neg, y0 = de5$pos, angle = 90, code = 3, 
       length = 0.05 )

# Adding a box to this panel

box()


## Plotting a density histogram for the elevation "low" in the inset


hist(de$height[de$elevation== "high"], main = NA,col = "lightblue",axes = F, freq = F,las = 1 )
densitys <- density(de$height[de$elevation== "high"])
lines(densitys)


##----------------------------------------##-------------------------------------------------##

## middle panel ##


#  data subsetting ; storing "mid" by ignoring high and low factor levels 


de2 = de[!de$elevation %in% c("high","low"), ]
de2 <- droplevels(de2) # remove unwanted factor levels
levels(de2$elevation) # generating "mid"


de3 <- group_by(.data = de2, spec ,treat ) %>%
  summarise(vertical_mean = mean(height, na.rm = TRUE),
            neg = mean(height, na.rm = TRUE) - se(height, na.rm = TRUE),
            pos = mean(height, na.rm = TRUE) + se(height, na.rm = TRUE))
de3

# middle panel Margin

par(mar = c(.24, 2.1, .95, .55))


format_wide <- acast(de3, treat ~ spec, value.var = "vertical_mean") # install.packages(reshape2) | libaray('reshape2')

# Creating a barplot for the elevation "mid"

barplot= barplot(format_wide, col = c("white", "blue"), ylim = c(0, 60), axisnames = F, beside = T
                 , axes = T) # barplot keeps the midpoints bars coordinates; creating texts/error bars 
# Panel box
box()

# Mid text 

mtext("Mid", side = 3, line = -1.7, cex = 0.8, outer = F)# Wide format data transformation

##--------- ---------##------ ------##
## Extra 

mtext("Height (cm)",  line = 2, cex = 0.6, side = 2)

##--------- ---------##------ ------##

# Creating error bars

arrows(x0 = barplot, x1 = barplot, y0= de3$pos, y1 = de3$neg , angle = 90, code = 3, length = 0.05)

## Low elevation Density histogram 

hist(de$height[de$elevation== "mid"], col = "lightblue", main = NA, las =
1, axes = F, freq = F)
densitys <- density(de$height[de$elevation== "mid"])
lines(densitys)




##----------------------------------------##-------------------------------------------------##
# bottom  panel


# Low elevation   

l = de[!de$elevation %in% c("high", "mid"), ] # "[]" allow data subsetting 

levels(l$elevation) # generating Low
l= droplevels(l) #  unexploited factor levels are dropped

# species, summary, standard error, treatment in  subsets generate  object group data.

library(dplyr) #  data summary package loaded via library ()

de1= group_by(.data = l, treat, spec) %>%
  summarise(vertical_mean = mean(height, na.rm = TRUE),
            neg = mean(height, na.rm = TRUE) - se(height, na.rm = TRUE),
            pos = mean(height, na.rm = TRUE) + se(height, na.rm = TRUE))
de1= de1[order(de1$treat), ]
de1

# Wide format data transformation 

library(reshape2)

#343434343434format_wide = acast(se1, treat ~ spec, value.var = "vertical_mean")

# Margin control

par(mar = c(2.1, 2.1, .95, .55))



# Creating a low elevation barplot for three species

barplot= barplot(format_wide, beside = T, ylim = c(0, 60), axes = T, axisnames
                 = F, col = c("white", "blue")) # 'bp' stores the coordinates of the midpoints of the bars, which you can now use to add error bars or text.

# Top panel text

mtext("Low", outer = F, cex = 0.8,line = -1.7, side = 3, font = 1)

# Error bars

arrows(x0 = barplot, x1 = barplot, y0= de1$pos, y1 = de1$neg , length = 0.05, angle = 90, code = 3)


##--------- ---------##------ ------##
## Extra 

# species x-axis, displaying Podocarpus totara as P.totara, 
# Beilschmiedia tawa as B.tawa and Agathis australis as A.australis,
text(barplot, -2,
     c("A.australis","A.australis","B.tawa","B.tawa","P.totara","P.totara"
     ),
     srt = 45, adj = c(1, 0), xpd = NA, cex = 0.9)
##--------- ---------##------ ------##

#  box panel 
box()

hist(de$height[de$elevation== "low"], col = "lightblue", main = NA, las =
       2, axes = F, freq = F)


## low elevation density histogram 

densitys <- density(de$height[de$elevation== "low"])
lines(densitys)


```




## Part 3 ##


## Multli-panel map 



```{r}

# Packages  

library(ggplot2)
library(ggmap)
library(ggthemes) 
library(RStoolbox) 
library(raster) 
library(ggsn) 
library(mapdata) 
library(broom)
library(rmapshaper)
library(colorRamps)
library(RColorBrewer)

#install.packages(c('ggmap','ggthemes','ggplot2', 'ggmap', 'ggthemes', 'RStoolbox', 'ggsn', 'raster', 'mapdata', 'broom', 'rmapshaper', 'colorRamps','RColorBrewer'))

#install.packages(c('RStoolbox', 'ggsn', 'mapdata', 'rmapshaper', 'colorRamps',type = "binary"))

```


Australia States 

```{r}
# Australia States

aus = getData(name = "GADM", country= "Australia", level= 1)
nrow(tidy(aus)) # 1426114 polygon areas


# Adding Australia State colours

cols= colorRampPalette(colors= c("red", "orange", "grey", "darkgreen", "purple")) 

# creating a Australia State map with distinct colours

t = ggplot(data= aus, aes(x= long, y= lat,map_id= id, group= group)) + 
  theme_map() +
  coord_quickmap() +
  geom_polygon(size= 0.5,aes(fill= id)) + 
  guides(fill= F) +
  scale_fill_manual(values = cols(n= 11)) 
  t

#  creating labels or texts of states in the Australia map 

states= t + annotate(geom = "text", x = 122, y = -24.8485, label = "Western \n Australia") +
  annotate(geom = "text", x = 133.5, y = -19.9485, label = "Northern\nTerritory") + 
  annotate(geom = "text", x = 134.5, y = -28.9485, label = "South \n Australia") + 
  annotate(geom = "text", x = 144.5, y = -23, label = "Queensland") + 
  annotate(geom = "text", x = 147.5, y = -32, label = "New South \n Wales") +
  annotate(geom = "text", x = 144.8, y = -36.9, label = "Victoria") +
  annotate(geom = "text", x = 146.8, y = -42, label = "Tasmania") 

states
```



Digital elevation models (DEM) of Australia 

```{r}
## Download altitude dataset
oz_alt= getData(name = "alt", country = "AUS") 

# selects australia via country = "AUS" and elevation from name = "alt" 
oz_alt
# Raster with elevation data can calculate the aspect, 
#any remaining terrain traits and slope of an area 
# (note: coordinate reference system (lon/lat) need to be in meters).

oz_terr= terrain(opt = c("slope", "aspect"), x = oz_alt[[1]])

oz_terr$slope # indiates australia geograpical information 

## Compute the hill shade from the slope and aspect layer 

oz_hillshade= hillShade(slope = oz_terr$slope, aspect = oz_terr$aspect)

## DEM map of NZ

## Crop the map by finding the studied site


s= extent(c(xmin =  86.842428, xmax = 158.068268, ymin = -49.887759, ymax =-10.240108)) # ss is the studied site

## the hill shade raster is cropped via reference ss ( studied site)

oz_hillshade2= crop(x = oz_hillshade, y = s)

## Same for the elevations

oz_alt2= crop(x = oz_alt, y = s)

## Colour palette for the DEM

library(RColorBrewer)


bbg <- colorRampPalette(colors = brewer.pal(n = 11, name = "BrBG")) 


RdYlBu <- colorRampPalette(colors = rev(brewer.pal(n = 11, "RdYlBu")))

# creating a DEM (note use help file or tab to find the arguments functions, values, usage, details)

oz.elv=  ggR(oz_hillshade2) + ggR(oz_alt2, geom_raster = T, ggLayer = T, alpha = 0.75, ggObj = T) + 
  
         
  
  scale_fill_gradientn(colours = bbg(n = 1500), 
                       
                       
                      na.value = "transparent", limits = c(0, 1500), name = "Elevation", 
                       
                        expand = c(2, 1),
                       
                             guide = guide_colourbar(nbin = 2000, barwidth = 0.7, 
                                               
                                    barheight = 10, 
                                               
                                    title.theme = element_text(size = 12), 
                                               
                                    label.theme = element_text(size = 10), 
                                               
                                    draw.ulim = F)) + 
                                              
                                    coord_quickmap()  + 
                                                    
                                    theme_map() + 
  
  
theme(legend.position = c(0.9, 0.4),legend.background = element_rect(colour = NA, fill = NA))

oz.elv
```



Adding the maps together 


```{r}
#install.packages('cowplot')
library(cowplot)
library(gridExtra)# generates grid.arrange() function
library(cowplot)
library(gridExtra)
```


```{r}
grid.arrange (oz.elv,states, layout_matrix= 
                rbind(c(2, 2, 1, 1),
                     c(2, 2, 1, 1),
                      c(2, 2, 1, 1),
                      c(2, 2, 1, 1)),
              heights = c(10, 10, 0, 0), widths = c(1, 1, 2, 2))
```


